---
title: "Planned Giving Donor Model Comparison"
author: "Kevin Kusuma"
date: "6/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
library(janitor)
library(ROSE)
library(smotefamily)
library(caTools)
library(DomoR)

data_ori <- read_csv("PG 20180630 Merged with EASI_Ready for Analysis CSV.csv")
# Data Filters Applied
data_ready <- data_ori %>% 
  filter(`Current Record Status`!='D', Age>=40,
         `Total U gifts`>1, 
         `Household Rating`!=0, 
         `Record Type 1 (Primary)`!= 'Student', 
         `Record Type 1 (Primary)`!= 'Former Student',
         `Latest U Gift Age`>=18, 
         is.na(`Household Rating`)==FALSE) %>% 
  clean_names("snake")

data_kev <- data_ready %>% 
  mutate(pg_donor = factor(ifelse(pg_donor==1,'Yes','No'), levels = c('Yes','No')),
         age_group = factor(floor(age/10)*10),
         address_sl25 = factor(address_sl25),
         has_attended_student_act = factor(has_attended_student_act),
         revieved_any_awards_honors = factor(revieved_any_awards_honors),
         former_uu_payroll_deduction = factor(former_uu_payroll_deduction),
         sports_participation = factor(sports_participation),
         greek_participation = factor(greek_participation),
         has_made_phonathon_pledge = factor(has_made_phonathon_pledge),
         presidents_club_current = factor(presidents_club_current),
         presidents_club_former = factor(presidents_club_former),
         crimson_club_member_current = factor(crimson_club_member_current),
         crimson_club_member_ever = factor(crimson_club_member_ever),
         any_committee_participation = factor(any_committee_participation),
         gender= factor(gender),
         has_attended_activities = factor(has_attended_activities),
         has_multiple_properties = factor(has_multiple_properties),
         multiple_u_degrees = factor(multiple_u_degrees),
         undergrad_degree = factor(undergrad_degree),
         has_any_u_contact_reports = factor(has_any_u_contact_reports),
         project_management_team = factor(project_management_team),
         marital_status_description = factor(marital_status_description),
         both_entity_and_spouse_u_alum = factor(both_entity_and_spouse_u_alum),
         has_direct_family_u_alum = factor(has_direct_family_u_alum),
         latest_u_gift_amount_group = floor(latest_u_gift_amount/100)*100,
         uu_years_of_giving_group = floor(uu_years_of_giving/10)*10,
         number_of_u_degrees_group = floor(number_of_u_degrees/10)*10,
         num_of_activities_attended_group = floor(num_of_activities_attended/10)*10,
         total_u_gifts_group = floor(total_u_gifts/100)*100,
         average_u_gift_5_yr_group = floor(average_u_gift_5_yr/100)*100,
         latest_u_gift_age_group = floor(latest_u_gift_age/10)*10,
         average_u_gift_group = floor(average_u_gift/100)*100,
         years_latest_u_gift_group = floor(years_latest_u_gift/10)*10,
         years_earliest_u_gift_group = floor(years_earliest_u_gift/10)*10,
         number_of_children,
         number_u_contact_reports_5_yr,
         num_of_acts_attended_5_yr,
         has_any_gifts_by_credit_card = factor(has_any_gifts_by_credit_card),
         has_phone_handling = factor(has_phone_handling),
         parent_of_u_alum = factor(parent_of_u_alum),
         child_of_u_alum = factor(child_of_u_alum),
         total_u_gift_count_group = floor(total_u_gift_count/10)*10,
         total_u_gift_count_5_yr_group = floor(total_u_gift_count_5_yr/10)*10) %>%
  select(pg_donor,
         average_u_gift_group,
         years_latest_u_gift_group,
         years_earliest_u_gift_group,
         number_of_children,
         number_u_contact_reports_5_yr,
         num_of_acts_attended_5_yr,
         has_any_gifts_by_credit_card,
         has_phone_handling,
         latest_u_gift_age_group,
         has_attended_student_act,
         revieved_any_awards_honors,
         former_uu_payroll_deduction,
         sports_participation,
         greek_participation,
         has_made_phonathon_pledge,
         has_attended_activities,
         crimson_club_member_current,
         crimson_club_member_ever,
         presidents_club_current,
         presidents_club_former,
         any_committee_participation,
         project_management_team,
         age_group,
         any_mailing_list,
         marital_status_description,
         both_entity_and_spouse_u_alum,
         household_rating,
         has_any_u_contact_reports,
         number_u_contact_reports,
         uu_years_of_giving_group,
         has_direct_family_u_alum,
         gender,
         latest_u_gift_amount_group,
         average_u_gift_5_yr_group,
         record_type_1_primary,
         degree_school_1,
         number_of_u_degrees_group,
         num_of_activities_attended_group,
         total_u_gifts_group,
         has_children,
         address_sl25,
         degree_department_1,
         wpot_rating,
         undergrad_degree,
         masters_degree,
         doctorate_degree,
         has_multiple_properties,
         has_direct_family_u_alum,
         total_u_gift_count_group,
         total_u_gift_count_5_yr_group,
         num_of_student_acts_attended,
         number_u_contact_reports_5_yr,
         number_u_contact_reports_10_yr) %>% 
  na.omit()

data_kev_1 <- dummyVars(pg_donor~., data = data_kev, fullRank = T) %>% 
  predict(newdata=data_kev)

data_kev_2 <- data_kev_1[ , -nzv(data_kev_1)]

data_kev_nzv <- data_kev_1[ , nzv(data_kev_1)]

data_kev_df1 <- data.frame(data_kev_2)
data_kev_df2 <- data.frame(data_kev_nzv)

data_kev_df1$pg_donor <- data_kev$pg_donor
data_kev_df1$project_management_team.1 <- data_kev_df2$project_management_team.1
data_kev_df1$num_of_activities_attended_group <- data_kev_df2$num_of_activities_attended_group
data_kev_df1 <- data_kev_df1 %>% 
  select(-degree_department_1Law)

# UOS
set.seed(212)
partition_kev <- createDataPartition(data_kev_df1$pg_donor, p=0.7, list = F)
train_kev <- data_kev_df1[partition_kev, ]
test_kev <- data_kev_df1[-partition_kev, ]

n_new <- train_kev %>% 
  filter(pg_donor=='No') %>% 
  nrow()

set.seed(212)
UOS <- ovun.sample(pg_donor~.,
                           data = train_kev,
                           method = "both",
                           N = n_new,
                           p = 0.5,
                           seed = 212)
uos_data <- UOS$data

# SMOTE
set.seed(212)
UOS_U <- ovun.sample(pg_donor~.,
                           data = train_kev,
                           method = "under",
                           N = 1650,
                           seed = 212)
uos_U <- UOS_U$data

set.seed(212)
smote_result <- SMOTE(X = uos_U[,-58],
                      target = uos_U$pg_donor,
                      K = 4,
                      dup_size = 9)
smote_data = smote_result$data %>% 
  mutate(pg_donor = factor(class, levels = c('Yes','No'))) %>% 
  select(-class)

set.seed(212)
logistic_uos <- train(pg_donor~.,
                        uos_data,
                        method = "glm",
                        metric = "Sens",
                        trControl = trainControl(
                          method = "cv",
                          number = 10,
                          summaryFunction = twoClassSummary,
                          classProbs = T,
                          verboseIter = T),
                        preProcess= c("center", "scale"))
pred_log <- predict(logistic_uos, test_kev, type = "prob")

set.seed(212)
logistic_smt <- train(pg_donor~.,
                        smote_data,
                        method = "glm",
                        metric = "Sens",
                        trControl = trainControl(
                          method = "cv",
                          number = 10,
                          summaryFunction = twoClassSummary,
                          classProbs = T,
                          verboseIter = T),
                        preProcess= c("center", "scale"))
pred_log_smt <- predict(logistic_smt, test_kev, type = "prob")

rf_tuning <- data.frame(mtry = 28,
                        splitrule = "extratrees",
                        min.node.size = 10)

set.seed(212)
rf_uos <- train(pg_donor~.,
                  uos_data,
                  method = "ranger",
                  metric= "Sens",
                  tuneGrid = rf_tuning,
                  importance = "permutation",
                  trControl = trainControl(
                    method = "cv",
                    number = 10,
                    summaryFunction = twoClassSummary,
                    classProbs = T,
                    verboseIter = T),
                  preProcess = c('center','scale'))
pred_rf <- predict(rf_uos, test_kev, type = "prob")

rf_tuning_smt <- data.frame(mtry = 2,
                        splitrule = "extratrees",
                        min.node.size = 10)
set.seed(212)
rf_smt <- train(pg_donor~.,
                  smote_data,
                  method = "ranger",
                  metric= "Sens",
                  tuneGrid = rf_tuning_smt,
                  importance = "permutation",
                  trControl = trainControl(
                    method = "cv",
                    number = 10,
                    summaryFunction = twoClassSummary,
                    classProbs = T,
                    verboseIter = T),
                  preProcess = c('center','scale'))
pred_rf_smt <- predict(rf_smt, test_kev, type = "prob")

gbmGrid <-  expand.grid(interaction.depth = 5, 
                        n.trees = 1500, 
                        shrinkage = 0.1,
                        n.minobsinnode = 20)

set.seed(212)
gbm_uos <- train(pg_donor~.,
                  uos_data,
                  method = "gbm",
                  metric= "Sens",
                  tuneGrid = gbmGrid,
                  trControl = trainControl(
                    method = "cv",
                    number = 10,
                    summaryFunction = twoClassSummary,
                    classProbs = T,
                    verboseIter = T),
                  preProcess = c('center','scale'))
pred_gbm <- predict(gbm_uos, test_kev, type = "prob")

gbmGrid_smt <-  expand.grid(interaction.depth = 3, 
                        n.trees = 50, 
                        shrinkage = 0.1,
                        n.minobsinnode = 10)

set.seed(212)
gbm_smt <- train(pg_donor~.,
                  smote_data,
                  method = "gbm",
                  metric= "Sens",
                  tuneGrid = gbmGrid_smt,
                  trControl = trainControl(
                    method = "cv",
                    number = 10,
                    summaryFunction = twoClassSummary,
                    classProbs = T,
                    verboseIter = T),
                  preProcess = c('center','scale'))
pred_gbm_smt <- predict(gbm_smt, test_kev, type = "prob")

data_2017 <- fetch('5b6770ae-ca9c-40da-9b49-7288d6cb8729')
data_2017 <- data_2017 %>% 
  clean_names("snake") 

data_2017_ready <- data_2017 %>%
  mutate(pg_donor = factor(ifelse(pg_donor==1,'Yes','No'), levels = c('Yes','No')),
         age_group = factor(floor(as.numeric(ifelse(age=='NA',NA,age))/10)*10),
         address_sl25 = factor(address_sl25),
         has_attended_student_act = factor(has_attended_student_act),
         revieved_any_awards_honors = factor(revieved_any_awards_honors),
         former_uu_payroll_deduction = factor(former_uu_payroll_deduction),
         sports_participation = factor(sports_participation),
         greek_participation = factor(greek_participation),
         has_made_phonathon_pledge = factor(has_made_phonathon_pledge),
         presidents_club_current = factor(president_s_club_current),
         presidents_club_former = factor(president_s_club_former),
         crimson_club_member_current = factor(crimson_club_member_current),
         crimson_club_member_ever = factor(crimson_club_member_ever),
         any_committee_participation = factor(any_committee_participation),
         gender= factor(gender),
         household_rating = as.numeric(ifelse(household_rating=='NA',NA,household_rating)),
         has_attended_activities = factor(has_attended_activities),
         has_multiple_properties = factor(has_multiple_properties),
         multiple_u_degrees = factor(multiple_u_degrees),
         undergrad_degree = factor(undergrad_degree),
         has_any_u_contact_reports = factor(has_any_u_contact_reports),
         project_management_team = factor(project_management_team),
         marital_status_description = factor(marital_status_description),
         both_entity_and_spouse_u_alum = factor(both_entity_and_spouse_u_alum),
         has_direct_family_u_alum = factor(has_direct_family_u_alum),
         latest_u_gift_amount_group = floor(latest_u_gift_amount/100)*100,
         uu_years_of_giving_group = floor(uu_years_of_giving/10)*10,
         number_of_u_degrees_group = floor(number_of_u_degrees/10)*10,
         num_of_activities_attended_group = floor(num_of_activities_attended/10)*10,
         total_u_gifts_group = floor(total_u_gifts/100)*100,
         average_u_gift_5_yr_group = floor(average_u_gift_5_yr/100)*100,
         latest_u_gift_age_group = floor(as.numeric(ifelse(latest_u_gift_age=='NA',NA,latest_u_gift_age))/10)*10,
         average_u_gift_group = floor(average_u_gift/100)*100,
         years_latest_u_gift_group = floor(as.numeric(ifelse(years_latest_u_gift=='NA',NA,years_latest_u_gift))/10)*10,
         years_earliest_u_gift_group = floor(as.numeric(ifelse(years_earliest_u_gift=='NA',NA,years_earliest_u_gift))/10)*10,
         number_of_children,
         number_u_contact_reports_5_yr,
         num_of_acts_attended_5_yr,
         has_any_gifts_by_credit_card = factor(has_any_gifts_by_credit_card),
         has_phone_handling = factor(has_phone_handling),
         parent_of_u_alum = factor(parent_of_u_alum),
         child_of_u_alum = factor(child_of_u_alum),
         total_u_gift_count_group = floor(total_u_gift_count/10)*10,
         total_u_gift_count_5_yr_group = floor(total_u_gift_count_5_yr/10)*10) %>% 
  select(pg_donor,
         average_u_gift_group,
         years_latest_u_gift_group,
         years_earliest_u_gift_group,
         number_of_children,
         number_u_contact_reports_5_yr,
         num_of_acts_attended_5_yr,
         has_any_gifts_by_credit_card,
         has_phone_handling,
         latest_u_gift_age_group,
         has_attended_student_act,
         revieved_any_awards_honors,
         former_uu_payroll_deduction,
         sports_participation,
         greek_participation,
         has_made_phonathon_pledge,
         has_attended_activities,
         crimson_club_member_current,
         crimson_club_member_ever,
         presidents_club_current,
         presidents_club_former,
         any_committee_participation,
         project_management_team,
         age_group,
         any_mailing_list,
         marital_status_description,
         both_entity_and_spouse_u_alum,
         household_rating,
         has_any_u_contact_reports,
         number_u_contact_reports,
         uu_years_of_giving_group,
         has_direct_family_u_alum,
         gender,
         latest_u_gift_amount_group,
         average_u_gift_5_yr_group,
         record_type_1_primary,
         degree_school_1,
         number_of_u_degrees_group,
         num_of_activities_attended_group,
         total_u_gifts_group,
         has_children,
         address_sl25,
         degree_department_1,
         wpot_rating,
         undergrad_degree,
         masters_degree,
         doctorate_degree,
         has_multiple_properties,
         has_direct_family_u_alum,
         total_u_gift_count_group,
         total_u_gift_count_5_yr_group,
         num_of_student_acts_attended,
         number_u_contact_reports_5_yr,
         number_u_contact_reports_10_yr)

data_2017_feed <- dummyVars(pg_donor~., data_2017_ready, fullRank = T) %>% 
  predict(newdata=data_2017_ready)

data_2017_feed <- preProcess(data_2017_feed, "medianImpute") %>% 
  predict(newdata=data_2017_feed)

data_2017_feed_a <- data.frame(data_2017_feed)

data_2017_feed <- data_2017_feed[ ,-nzv(data_2017_feed)]

data_2017_feed <- data.frame(data_2017_feed)

data_2017_feed$pg_donor <- data_2017_ready$pg_donor
data_2017_feed$number_of_children <- data_2017_feed_a$number_of_children
data_2017_feed$number_u_contact_reports_5_yr <- data_2017_feed_a$number_u_contact_reports_5_yr
data_2017_feed$greek_participation.1 <- data_2017_feed_a$greek_participation.1
data_2017_feed$crimson_club_member_current.1 <- data_2017_feed_a$crimson_club_member_current.1
data_2017_feed$crimson_club_member_ever.1 <- data_2017_feed_a$crimson_club_member_ever.1
data_2017_feed$presidents_club_former.1 <- data_2017_feed_a$presidents_club_former.1
data_2017_feed$any_committee_participation.1 <- data_2017_feed_a$any_committee_participation.1
data_2017_feed$marital_status_description.Married <- data_2017_feed_a$marital_status_description.Married
data_2017_feed$marital_status_description.Widowed <- data_2017_feed_a$marital_status_description.Widowed
data_2017_feed$household_rating <- data_2017_ready$household_rating
data_2017_feed$has_any_u_contact_reports.1 <- data_2017_feed_a$has_any_u_contact_reports.1
data_2017_feed$uu_years_of_giving_group <- data_2017_feed_a$uu_years_of_giving_group
data_2017_feed$average_u_gift_5_yr_group <- data_2017_feed_a$average_u_gift_5_yr_group
data_2017_feed$degree_school_1Education <- data_2017_feed_a$degree_school_1Education
data_2017_feed$degree_school_1Engineering <- data_2017_feed_a$degree_school_1Engineering
data_2017_feed$degree_school_1Law <- data_2017_feed_a$degree_school_1Law
data_2017_feed$degree_school_1Medicine <- data_2017_feed_a$degree_school_1Medicine
data_2017_feed$degree_school_1Science <- data_2017_feed_a$degree_school_1Science
data_2017_feed$degree_department_1Communication <- data_2017_feed_a$degree_department_1Communication
data_2017_feed$degree_department_1Teaching.and.Learning <- data_2017_feed_a$degree_department_1Teaching.and.Learning
data_2017_feed$doctorate_degree <- data_2017_feed_a$doctorate_degree
data_2017_feed$has_multiple_properties.1 <- data_2017_feed_a$has_multiple_properties.1
data_2017_feed$total_u_gift_count_group <- data_2017_feed_a$total_u_gift_count_group
data_2017_feed$total_u_gift_count_5_yr_group <- data_2017_feed_a$total_u_gift_count_5_yr_group
data_2017_feed$number_u_contact_reports_10_yr <- data_2017_feed_a$number_u_contact_reports_10_yr
data_2017_feed$project_management_team.1 <- data_2017_feed_a$project_management_team.1
data_2017_feed$num_of_activities_attended_group <- data_2017_feed_a$num_of_activities_attended_group
data_2017_feed$number_u_contact_reports <- data_2017_feed_a$number_u_contact_reports
glimpse(data_2017_feed)
table(train_kev$household_rating)
table(data_2017_feed$household_rating)
data_2017_feed_b <- data_2017_feed %>% 
  filter(household_rating>0)
```

## Extreme Imbalanced Class Proportion of PG Donor in 2018 dataset

# Extremely Imbalanced Dataset vs Machine Learning Algorithms
Because the number of Planned Giving donors in the 2018 dataset is extremely low (less than 1%) compared to the overall dataset and most Machine Learning Algorithms (Logistic, RF, GBM, etc) assume that the class proportion is relatively balanced, it is extremely necessary to balance the training dataset before building a model. Here we compare two popular methods/ techniques to balance the dataset, namely, UOS (Under/Over Sampling) & SMOTE (Sythetic Minority Oversampling Technique) before building predictive models.

# UOS (Under/Over Sampling)
As the name suggests, we balance the class proportion by reducing the number of the majority class (Non PG) through undersampling and increasing the number of the minority (PG) through oversampling. We use ROSE package to complete this task.

# SMOTE (Synthetic Minority Oversampling Technique)
This method is oversampling the minority class using KNN (K-Nearest Neighbor) principle. This is considered a more advanced technique than the UOS.

# Machine Learning Models
We use three different types of Machine Learning Algorithms, namely: Logistic Regression, Random Forest, and Gradient Boosting Machine. Because we will train each algorithm with 2 different training datasets (One from SMOTE, the other from UOS), we will have 6 different models. 

# Creating Training Datasets and Training the Models
I have completed all the necessary steps (Creating the datasets & Training the Models) so that we can now assess the performances of the models. 

## Using 2017 Dataset to Measure the Models' Performances
We use 2017 dataset as the Test dataset because it's a fresh dataset that the models have not seen and we already know the true labels of the PG Donor for each observation (row) in the dataset.

# Logistic Regression on UOS Training Dataset
```{r}
# Logistic UOS
pred_loguos_2017 <- predict(logistic_uos, data_2017_feed_b, type = "prob")
colAUC(pred_loguos_2017, data_2017_feed_b$pg_donor, plotROC = T)
```

# Logistic Regression on SMOTE Training Dataset
```{r}
# Logistic SMOTE
pred_logsmt_2017 <- predict(logistic_smt, data_2017_feed_b, type = "prob")
colAUC(pred_logsmt_2017, data_2017_feed_b$pg_donor, plotROC = T)
```

# Random Forest on UOS Dataset
```{r}
# Random Forest UOS
pred_rfuos_2017 <- predict(rf_uos, data_2017_feed_b, type = "prob")
colAUC(pred_rfuos_2017, data_2017_feed_b$pg_donor, plotROC = T)
```

# Random Forest on SMOTE Dataset
```{r}
# Random Forest SMOTE
pred_rfsmt_2017 <- predict(rf_smt, data_2017_feed_b, type = "prob")
colAUC(pred_rfsmt_2017, data_2017_feed_b$pg_donor, plotROC = T)
```

# Gradient Boosting Machine on UOS Dataset
```{r}
# GBM UOS
pred_gbmuos_2017 <- predict(gbm_uos, data_2017_feed_b, type = "prob")
colAUC(pred_gbmuos_2017, data_2017_feed_b$pg_donor, plotROC = T)
```

# Gradient Boosting Machine on SMOTE Dataset
```{r}
# GBM SMOTE
pred_gbmsmt_2017 <- predict(gbm_smt, data_2017_feed_b, type = "prob")
colAUC(pred_gbmsmt_2017, data_2017_feed_b$pg_donor, plotROC = T)
```

## Conclusion
Of the 6 models above, the Gradient Boosting Machine (GBM) model trained using SMOTE technique has the highest ROC. Thus, it is best model we currently have on this documentation.

## Appendix
# Most Important Variables in the GBM model with SMOTE:
```{r}
summary(gbm_smt)
```
# Summary of the Logistic Model Using SMOTE
```{r}
summary(logistic_uos)
varImp(logistic_uos)
```